BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-03-19 paper/paper.qmd
multi scale section

Then for any row $i\in[I]$, the sum of the row is $\epsilon\sqrt{J}$ close to 1

@@

We propose a multiresolution approach inspired by multi-grid [] and wavelets [] that factorizes the data at progressively finer scales, using the . The can greatly speed up how fast large scale tensors can be factorized.

@@

A natural extension of the example shown in @sec-basic-multi-scale would be to higher dimensional distributions. For example, we could consider an order-$3$ tensor where the horizontal slices correspond to a $2$-dimensional discretization of a bivariate density. Entries of the input tensor $Y$ would be given by $Y[i,j,k] = f_i(x_j, y_k)$ for continuous probability density functions $f_i:\mathbb{R}^2\rightarrow \mathbb{R}_+$ for a 2D grid of points $(x_j, y_k)$. In this example, the second and third dimensions would be continuous.

Factors that de

, and $J_s$ would be a discretization that uses every $2^{s}$ points, or $J_s = 2^{S-s}+1$ points in total.

The basic approach will be to factorize the coarsest version of $Y$, $Y_S\in\mathbb{R}^{I\times J_S}$, to obtain $A_S$ and $B_S$.

>>>>

Then for any row $i\in[I]$, the sum of the row $A[i,:]$ is $\epsilon\sqrt{J}$ close to 1

@@

We propose a multiresolution approach inspired wavelets [@benedetto_wavelets_1993] and multigrid [@trottenberg_multigrid_2001] that factorizes the data at progressively finer scales. This can greatly speed up how fast large scale tensors can be factorized.

@@

An example of this setting would be an extension of the example shown in @sec-basic-multi-scale to higher dimensional distributions. We could consider an order-$3$ tensor where the horizontal slices correspond to a $2$-dimensional discretization of a bivariate density. Entries of the input tensor $Y$ would be given by $Y[i,j,k] = f_i(x_j, y_k)$ for continuous probability density functions $f_i:\mathbb{R}^2\rightarrow \mathbb{R}_+$ for a 2D grid of points $(x_j, y_k)$. In this example, the second and third dimensions would be continuous.

If we want to perform a rank-$(R_1,\dots,R_N)$ Tucker decomposition of $Y=\left\llbracket A_0; A_1, \dots, A_N\right\rrbracket$, we can initialize the factors with a very corse factorization $A_{n_m}^{s}\in\mathbb{R}^{J_{n_m}^{s} \times R_{n_m}}$ where $J^{s}$ would be a discretization that uses every $2^{s}$ points, or more accurately

$$
J_{n_m}^{s} = 2^{\max(S_m-s, 0)}+1
$$

points in total. We select $S_m$ so that at the finest scale $s=0$, we have $J_{n_m}^{0}=I_{n_m}$. The dimensions that do not represent continuous values will use the full sized factors $A_n^{s}\in\mathbb{R}^{J_n \times R_n}$ with $J_n = I_n$, and the core will remain $A_0^{s}\in\mathbb{R}^{R_1\times\dots\times R_N}$ at all scales $s$.^[This assumes the dimensions of $Y$ where $Y$ is continuous have been discretized as one more than a power of two. The same idea holds with some other discretization plan, but becomes more complicated to express notationally and keep track of the number of points at each scale.]

The aim is to fit the product of the factors $\left\llbracket A_0^{s}; A_1^{s}, \dots, A_N^{s}\right\rrbracket$ to a lower resolution version of $Y\in\mathbb{R}^{I_1 \times \dots \times I_N}$, namely $Y^s\in\mathbb{R}^{J_1^s \times \dots \times J_N^s}$, and use the result to initialize a finer version of the factors $A$. The code for this looks like the following.

```julia
function multiscale_factorize(Y; kwargs...)
    scales, kwargs = initialize_scales(Y, kwargs)
    coarsest_scale, finer_scales... = scales

    # Factorize Y at the coarsest scale
    Yₛ = coarsen(Y, coarsest_scale; kwargs...)
    decomposition, stats, kwargs = factorize(Yₛ; kwargs...)

    # Factorize Y at progressively finer scales
    for scale in finer_scales
        # Use an interpolated version of the coarse factorization
        # as the initialization
        decomposition = interpolate(decomposition, scale; kwargs...)
        kwargs[:decomposition] = decomposition

        Yₛ = coarsen(Y, scale; kwargs...)
        decomposition, stats, kwargs = factorize(Yₛ; kwargs...)
    end
    return decomposition, stats, kwargs
end
```

Straightforward subsampled coarsening and constant interpolating can be used for `coarsen` and `interpolate`, but more sophisticated methods can be used in principle. Since the final solve of `factorize` is on the original sized problem, the choice of coarsening and interpolating only influences the initialization used at this finest scale. Bellow are examples of the basic coarsening and interpolation methods.

```julia
coarsen(Y::AbstractArray, scale::Integer; dims=1:ndims(Y), kwargs...) =
    Y[(d in dims ? axis[begin:scale:end] : axis for (d, axis) in enumerate(axes(Y)))...]

function interpolate(Y, scale; dims=1:ndims(Y), kwargs...)
    Y = repeat(Y; inner=(d in dims ? scale : 1 for d in 1:ndims(Y)))

    # Chop the last slice of repeated dimensions
    # since we only interpolate between the values
    return Y[(d in dims ? axis[begin:end-scale+1] : axis for (d, axis) in enumerate(axes(Y)))...]
end
```
